---
title: "Titanic - Random Forest Model"
author: "Umer Bhutto"
date: "June 17, 2019"
output: html_document
---

score = 0.80382

submission = https://www.kaggle.com/c/titanic/overview


The sinking of the RMS Titanic is one of the most infamous shipwrecks in history.  On April 15, 1912, during her maiden voyage, the Titanic sank after colliding with an iceberg, killing 1502 out of 2224 passengers and crew. This sensational tragedy shocked the international community and led to better safety regulations for ships.

One of the reasons that the shipwreck led to such loss of life was that there were not enough lifeboats for the passengers and crew. Although there was some element of luck involved in surviving the sinking, some groups of people were more likely to survive than others, such as women, children, and the upper-class.

In this challenge, we ask you to complete the analysis of what sorts of people were likely to survive. In particular, we ask you to apply the tools of machine learning to predict which passengers survived the tragedy.

Survival - 0 = No | 1 = Yes
Pclass - Ticket class
sex - Sex of passenger
Age - Age in years
Sibsp - # of siblings / spouses aboard the Titanic
parch - # of parents / children aboard the Titanic
ticket - ticket number
fare - Passenger fare
cabin - Cabin number
embarked - Port of Embarkation (C = Cherbourg, Q = Queenstown, S = Southampton)

Notes:

pclass: A proxy for socio-economic status (SES)
1st = Upper
2nd = Middle
3rd = Lower

age: Age is fractional if less than 1. If the age is estimated, is it in the form of xx.5

sibsp: The dataset defines family relations in this way...
Sibling = brother, sister, stepbrother, stepsister
Spouse = husband, wife (mistresses and fiancés were ignored)

parch: The dataset defines family relations in this way...
Parent = mother, father
Child = daughter, son, stepdaughter, stepson
Some children travelled only with a nanny, therefore parch=0 for them.

## Read Data

The dataset is divided into a training and testing dataset. We will read in these datasets and attempt to combine them
```{r}

traindf <- read.csv("train.csv")
testdf <- read.csv("test.csv")

#understand the structure of the 2 datasets
str(traindf)
str(testdf)
# testdf is missing the Survived column
testdf$Survived <- NA

# Create a column that states whether the dataset is training or not
traindf$IsTrain <- TRUE
testdf$IsTrain <- FALSE

#combine the datasets and look at the summary and structure
data.full <- rbind(traindf, testdf)
str(data.full)
summary(data.full)
nrow(data.full)

# Survived - 418 NA Values
# Age - 263 NA values
# Fare - 1 NA
# Embarked - 2 missing
# Cabin - 1014 missing

for(i in 1:13){
  m <- data.full[is.na(data.full[,i]) | data.full[,i] == "",]
  print(m)
}

```


## Data Cleansing

We will attempt to change / manipulate the missing values as shown in the dataset
```{r}

library(dplyr)
library(ggplot2)

# Change some columns to factors
data.full$Survived <- as.factor(data.full$Survived)
data.full$Pclass <- as.factor(data.full$Pclass)
data.full$Survived <- as.factor(data.full$Survived)
data.full$SibSp <- as.factor(data.full$SibSp)
data.full$Parch <- as.factor(data.full$Parch)

# Change Ticket to a numeric variable
data.full$Ticket <- as.numeric(data.full$Ticket)

# Change Name to character variable
data.full$Name <- as.character(data.full$Name)

# Survived - 418 NA Values -> This is due to the test df and are not important as we will predict this with our model later on

# Embarked - 2 missing values

data.full[data.full$Fare > 50,] %>% ggplot(aes(x = Fare, fill = Embarked)) + geom_histogram()
data.full[data.full$Fare > 50,] %>% ggplot(aes(x = Fare, fill = Sex)) + geom_histogram()
data.full[data.full$Fare >= 80,] %>% ggplot(aes(x = Sex, y = Embarked, color = )) + geom_jitter()


nrow(data.full[data.full$Embarked == "S" & data.full$Fare > 70 & data.full$Sex == "female",])
nrow(data.full[data.full$Embarked == "C" & data.full$Fare > 70 & data.full$Sex == "female",])

# These two graphs show that people in the 70 - 80 range mostly got in from C or S. Men and Women were about equal in this price range. There were about equal women who embarked on both C and S in this price range. We will choose C as there are more people who bought tickets at C when the fare > 70 and the passengers are women

data.full[data.full$Embarked == "","Embarked"] <- "C"

# Fare - 1 NA Value

data.full %>% ggplot(aes(x = Fare, fill = Pclass)) + geom_histogram()
data.full[data.full$Age > 60,] %>% ggplot(aes(x = Fare, fill = Pclass)) + geom_histogram()
# It seems that the Pclass = 3 is the lowest of the classes. As a result, we will take the mean of the values for ages 55 and up so that we have more data to work with for only P class 3.

m <- mean(data.full[data.full$Age > 55 & data.full$Pclass == 3,"Fare"], na.rm = T)
data.full[is.na(data.full$Fare), "Fare"] <- m

# Age - 263 NA values
data.full[!is.na(data.full$Age),] %>% ggplot(aes(x = Age, fill = Embarked)) + geom_histogram()

# We will first get the median age for each embarked destination and for each gender. This median will then be assigned to that specific group of NA values
embarked_vals <- c("C","Q","S")
sex_vals <- c("female","male")

for(j in sex_vals){
  for(i in embarked_vals){
    m <- median(data.full[data.full$Embarked == i & data.full$Sex == j,"Age"], na.rm = T)
    assign(paste(i,"_",j,sep=""), m)
}
}

#male values
data.full[data.full$Embarked == "C" & data.full$Sex == "male" & is.na(data.full$Age),"Age"] <- C_male
data.full[data.full$Embarked == "S" & data.full$Sex == "male" & is.na(data.full$Age),"Age"] <- S_male
data.full[data.full$Embarked == "Q" & data.full$Sex == "male" & is.na(data.full$Age),"Age"] <- Q_male

#female values
data.full[data.full$Embarked == "C" & data.full$Sex == "female" & is.na(data.full$Age),"Age"] <- C_female
data.full[data.full$Embarked == "S" & data.full$Sex == "female" & is.na(data.full$Age),"Age"] <- S_female
data.full[data.full$Embarked == "Q" & data.full$Sex == "female" & is.na(data.full$Age),"Age"] <- Q_female

#remove extra values
rm(C_female) & rm(C_male) & rm(S_female) & rm(S_male) & rm(Q_female) & rm(Q_male) & rm(i) & rm(j) & rm(m) & rm(sex_vals) & rm(embarked_vals) & rm(l)
```

We will look at transformation or additional features that can be made

```{r}
# Look at Fare variable
ggplot(data = data.full, aes(x = Fare)) + geom_histogram()
# Looks very left skewed, but there are many 0 values so we will instead focus on other things

# We will get the family name to create a variable to find the family size
data.full$Name <- as.character(data.full$Name)
for(i in 1:nrow(data.full)){
  data.full[i,"LastName"] <- substr(data.full[i,"Name"],1,regexpr(",",data.full[i,"Name"]) - 1)
}
data.full$LastName <- as.character(data.full$LastName)

# Create a family variable
# any people <= 2 is a small family
# any people > 2 is a large family
data.full$LastNameRpt <- 0
for(i in 1:nrow(data.full)){
  data.full[i,"LastNameRpt"]<-nrow(data.full[!is.na(match(data.full$LastName,data.full[i,"LastName"])),])
}

# Change family size to factor and plot the results
data.full$LastNameRpt <- as.factor(data.full$LastNameRpt)

ggplot(data = data.full[data.full$IsTrain == TRUE,], aes(x = LastNameRpt, fill = Survived)) + geom_bar()

# Look at titles of people
data.full$Title <- NA

for(i in 1:nrow(data.full)){
  data.full[i,"Title"] <- substr(data.full[i,"Name"],regexpr(",", data.full[i,"Name"]) + 2, 
                            regexpr("\\.",data.full[i,"Name"]) - 1)
}

data.full$Title <- as.factor(data.full$Title)
table(data.full$Title)

# Take a look at Cabin and see if we can fill in the values
data.full$Deck <- NA
for(i in 1:nrow(data.full)){
  data.full[i,"Deck"] <- substr(data.full[i,"Cabin"],1,1)
}
data.full[data.full$Deck == "","Deck"] <- "U"
data.full$Deck <- as.factor(data.full$Deck)
# We will not use this as it doesn't have good predictive power
data.full$Deck <- NULL
data.full$Cabin <- NULL

# combine SibSp and Parch to create family size
data.full$Family.Size <- as.numeric(data.full$SibSp) + as.numeric(data.full$Parch) + 1
data.full$Family.Size <- as.factor(data.full$Family.Size)

```

## Data Visualization

We now try to analyze the most important variables and their affect on the dependent variable
```{r}

library(gridExtra)

data.full[!is.na(data.full$Survived),] %>% ggplot(aes(x = Embarked, y = Survived, color = Sex)) + geom_jitter()
#As seen here the majority of people that survived were from "S" and "C"

data.full[!is.na(data.full$Survived),] %>% ggplot(aes(x = Survived, y = Pclass, color = Sex)) + geom_jitter()
# Most from Pclass 3 did not survive

data.full[!is.na(data.full$Survived),] %>% ggplot(aes(x = Survived, fill = Sex)) + geom_bar()
# Most people that survived were female. 

data.full[!is.na(data.full$Survived),] %>% ggplot(aes(x = Survived, fill = Pclass)) + geom_bar()
# most people that survived were from Pclass 1. Most people that did not survive were from Pclass 3

data.full[!is.na(data.full$Survived),] %>% ggplot(aes(x = Survived, y = Fare)) + geom_jitter() + geom_boxplot(alpha = 0.3)
# there doesn't seem to be much of a relationship between fare and survival

data.full[!is.na(data.full$Survived),] %>% ggplot(aes(x = Survived, y = Age)) + geom_jitter() + geom_boxplot(alpha = 0.3)
# there doesn't seem too much of a correlation between survival and age

data.full[!is.na(data.full$Survived),] %>% ggplot(aes(x = Survived, fill = SibSp)) + geom_bar()
# Doesn't seem like there is too much of a correlation

str(data.full)

```

## Data Partition

Divide the dataset back into the training and testing dataset

```{r}

set.seed(123)
traindf <- data.full[data.full$IsTrain == TRUE,]
testdf <- data.full[data.full$IsTrain == FALSE,]

#check
nrow(data.full) == nrow(traindf) + nrow(testdf)
str(traindf)

```

## Random Forest

Create the Random Forest

```{r}

form1 <- as.formula(Survived ~ Pclass + Sex + Age + SibSp + Parch + Ticket + Fare + Embarked + Title +
                      Family.Size + LastNameRpt)

library(randomForest)
model.rf <- randomForest(formula = form1, 
                         data = traindf, 
                         ntree = 500, 
                         mtry = 3, 
                         nodesize = 0.01 * nrow(traindf))

survived.var <- predict(model.rf, newdata = testdf)
```

## Output File

Create the submission CSV document

```{r}
#to submit on Kaggle, it needs 2 columns: PassengerID and Survived
PassengerId <- testdf$PassengerId
output.df <- as.data.frame(PassengerId)
output.df$Survived <- survived.var
write.csv(output.df, "titanic_kaggle_submission.csv",row.names = F)

# We submitted this and we got a score of 0.8038
```


